Finiteness in Cubical Type Theory

### 2020-05-31 sections/finiteness-predicates.tex
finished cardinal

Next, we will show that split enumerability is equivalent to the existence of a
split surjection from a finite prefix of the natural numbers:
\begin{lemma} \label{split-enum-is-split-surj}
  \begin{equation}
    \mathcal{E}!(A) \simeq \Sigma (n : \mathbb{N}) , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}!(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \in \mathit{xs}
    && \text{def.~\ref{split-enum-def} }(\mathcal{E}!)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{fib}_{\text{snd}(\mathit{xs})}(x)
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{sp-surj-eqn} (sp-surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{sp-surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \twoheadrightarrow ! \; A )
    && \text{eqn.~\ref{sp-surj-arrow-eqn} } (\twoheadrightarrow!)
  \end{align*}
\end{proof}
\begin{agdalisting}
In our formalisation, the proof is a single line:
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
The only step which isn't definitional equality is the reassociation of
\(\Sigma\).
\ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)

@@

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power,
so any closure proof we have on one can be transferred to the other.
In particular, it means that manifest Bishop finiteness is closed under
\(\Sigma\).
\subsubsection{From Manifest Bishop Finiteness to Equivalence}
We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
\emph{equivalence} with \(\mathbf{Fin}\).
\begin{lemma} \label{bishop-equiv}
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{B}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
    && \text{def.~\ref{bish-def} }(\mathcal{B})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
    && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{isEquiv}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \simeq A )
    && \text{eqn.~\ref{equiv-def} } (\simeq)
  \end{align*}
\end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}
\subsection{Cardinal Finiteness}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

The following two lemmas are proven in
\cite{yorgeyCombinatorialSpeciesLabelled2014} (Proposition 2.4.9 and 2.4.10,
respectively), in much the same way as we have done here.
Our contribution for this section is simply the formalisation.
\begin{lemma}
  Given a cardinally finite type, we can derive the type's cardinality, as well
  as a propositionally truncated proof of equivalence with \(\textbf{Fin}\)s of
  the same cardinality.
  \begin{equation}
    \mathcal{C}(A) \rightarrow \Sigma {(n : \mathbb{N})} , \lVert \textbf{Fin}(n) \simeq A \rVert
  \end{equation}
\end{lemma}
\begin{lemma} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{lemma}
\subsection{Manifest Enumerability}
\begin{definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) \coloneqq \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{definition}
As with manifest Bishop finiteness, the only difference with this type and split
enumerability is the membership proof: here we have propositionally truncated
it.
This has two effects.
First, it means that this proof represents a true surjection (rather than a
split surjection) from \(\mathbf{Fin}\).
\begin{lemma}
  \begin{equation}
    \mathcal{E}(A) \simeq \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \twoheadrightarrow A)
  \end{equation}
\end{lemma}
The proof for this lemma is similar in structure to
lemma~\ref{split-enum-is-split-surj} and lemma~\ref{bishop-equiv}.

>>>>

Next, we will show that split enumerability is equivalent to the existence of a
split surjection from a finite prefix of the natural numbers:
\begin{lemma} \label{split-enum-is-split-surj}
  \begin{equation}
    \mathcal{E}!(A) \simeq \Sigma (n : \mathbb{N}) , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}!(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \in \mathit{xs}
    && \text{def.~\ref{split-enum-def} }(\mathcal{E}!)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{fib}_{\text{snd}(\mathit{xs})}(x)
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{sp-surj-eqn} (sp-surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{sp-surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \twoheadrightarrow ! \; A )
    && \text{eqn.~\ref{sp-surj-arrow-eqn} } (\twoheadrightarrow!) \; \qedhere
  \end{align*}
\end{proof}
\begin{agdalisting}
In our formalisation, the proof is a single line:
\ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{is-split-inj}
The only step which isn't definitional equality is the reassociation of
\(\Sigma\).
\ExecuteMetaData[agda/Data/Sigma/Properties.tex]{reassoc}
\end{agdalisting}
(The simplicity of this proof, by the way, is why we preferred the
container-based definition of lists over the traditional one.)

@@

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power,
so any closure proof we have on one can be transferred to the other.
In particular, it means that manifest Bishop finiteness is closed under
\(\Sigma\).
\subsubsection{From Manifest Bishop Finiteness to Equivalence}
We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
\emph{equivalence} with \(\mathbf{Fin}\).
\begin{lemma} \label{bishop-equiv}
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{B}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
    && \text{def.~\ref{bish-def} }(\mathcal{B})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
    && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{isEquiv}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \simeq A )
    && \text{eqn.~\ref{equiv-def} } (\simeq) \; \qedhere
  \end{align*}
\end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}
\subsection{Cardinal Finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \begin{multicols}{2}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
  \end{multicols}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

The following two lemmas are proven in
\cite{yorgeyCombinatorialSpeciesLabelled2014} (Proposition 2.4.9 and 2.4.10,
respectively), in much the same way as we have done here.
Our contribution for this section is simply the formalisation.
\begin{lemma}
  Given a cardinally finite type, we can derive the type's cardinality, as well
  as a propositionally truncated proof of equivalence with \(\textbf{Fin}\)s of
  the same cardinality.
  \begin{equation}
    \mathcal{C}(A) \rightarrow \Sigma {(n : \mathbb{N})} , \lVert \textbf{Fin}(n) \simeq A \rVert
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a cardinally-finite type, with proof \(F : \mathcal{C}(A)\).
  Our task is to extract a natural number \(n : \mathbb{N}\) representing the
  cardinality of \(A\), and a propositionally-truncated proof that \(A\) is
  equivalent to \(\mathbf{Fin}(n)\).

  Extracting the second component of the pair is trivial, as it itself is
  truncated.
  We will now focus on extracting the cardinality.

  Without the propositional truncation, \(\text{fst}\) would suffice for this
  task.
  Given that the pair is hidden under the truncation, then, we need a way to
  convert a function \(f : A \rightarrow B\) to \(g : \lVert A \rVert
  \rightarrow B\).
  This is precisely what eliminator \ref{elim-prop-coh} gives us.
  For our case, we need to show the following:
  \begin{equation}
    \frac{(n : \mathbb{N}) \; \; \; (p : \mathbf{Fin}(n) \simeq A) \; \; \;
          (m : \mathbb{N}) \; \; \; (q : \mathbf{Fin}(m) \simeq A)
        }{
          n \equiv m
        }
  \end{equation}
  Immediately we can construct the following term:
  \begin{equation}
    \begin{alignat}{3}
      \mathbf{Fin}(n) & \simeq A && (p) \\
                      & \simeq \mathbf{Fin}(m) && (q)
    \end{alignat}
  \end{equation}
  Given univalence we have \(\mathbf{Fin}(n) \equiv \mathbf{Fin}(m)\),
  and the rest of our task is to prove:
  \begin{equation}
    \frac{\mathbf{Fin}(n) \equiv \mathbf{Fin}(m)}{n \equiv m}
  \end{equation}

  This is a well-known chestnut in dependently-typed programming, and one that
  has a surprisingly tricky and complex proof.
  We do not include it here, since it has already been explored elsewhere, but
  it is present in our formalisation.
\end{proof}
\begin{lemma} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{lemma}
\begin{proof}
  Since we can already derive decidable equality from a proof of manifest Bishop
  finiteness, it suffices to show that decidable equality is itself a
  proposition.
  \begin{equation}
    \text{isProp}(\Pi(x, y : A) , \mathbf{Dec}(x \equiv y))
  \end{equation}
  First, it is clear that \(x \equiv y\) is a proposition: since the type \(A\)
  has decidable equality, by Hedburg's theorem \todo{ref here} it is a set,
  meaning precisely that \(x \equiv y\) is a proposition.

  Secondly, we know that any decision over a proposition is itself a
  proposition.
  For any two terms \(x, y: \mathbf{Dec}(A)\) we cannot have the case that one
  is a yes decision and the other is no: from that we could derive \(\bot\).
  If both are no then they are both equal since \(A \rightarrow \bot\) is a
  proposition through function extensionality.
  And finally if both are yes then we know they must be equal because the type
  decided over is itself a proposition.

  Finally, since we know that \(\mathbf{Dec}(x \equiv y)\) is a proposition, we
  can derive that \(\Pi(x, y : A) , \mathbf{Dec}(x \equiv y)\) is a proposition
  (through function extensionality), proving our goal.
\end{proof}
\subsection{Manifest Enumerability}
\begin{definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) \coloneqq \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{definition}
As with manifest Bishop finiteness, the only difference with this type and split
enumerability is the membership proof: here we have propositionally truncated
it.
This has two effects.
First, it means that this proof represents a true surjection (rather than a
split surjection) from \(\mathbf{Fin}\).
\begin{lemma}
  \begin{equation}
    \mathcal{E}(A) \simeq \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \twoheadrightarrow A)
  \end{equation}
\end{lemma}
The proof for this lemma is similar in structure to
lemma~\ref{split-enum-is-split-surj} and lemma~\ref{bishop-equiv}.
